\texttt{thornado}-transport: IMEX schemes for two-moment neutrino transport respecting Fermi-Dirac statistics

### 2019-02-14 Documents/M1/Astronum_2018/sections/TimeIntegration.tex

Since the set of realizable moments is convex, convex-invariant schemes, which maps the initial value remains in its domain, can be used to design realizability-preserving methods for the two-moment model.
Ideally, the scheme should also be high-order accurate and work well in the asymptotic diffusion limit (characterized by frequent collisions and long time scales).  
The following discussion considers the construction of such convex-invariant schemes.

@@

If the IMEX scheme is GSA, $\vect{u}^{n+1} = \vect{u}^{(s)}$.  
Moreover, if $c_{ij},\tilde{c}_{ij}\ge0$ and $a_{ii}>0$, each stage in Eq.~\eqref{eq:imexStagesRewrite} is a convex combination of explicit Euler steps (with time step $\hat{c}_{ij}\dt$), followed by a implicit Euler step.  
Each of the explicit Euler steps has a time step condition that ensures its realizability given by $\hat{c}_{ij}\,\dt\leq\dx$; the CFL condition of the scheme.
Using results proved in~\cite{chu_etal_2018} and discussed in Section~\ref{se:SpatialDiscretization}, the IMEX scheme is convex-invariant and realizability-preserving for the two-moment model in Section~\ref{se:SpatialDiscretization} provided
\begin{equation}
  \max(\hat{c}_{ij})\,\dt \leq \dx.  
\end{equation}
(This CFL condition becomes more restrictive with high-order DG spatial discretization~\cite{chu_etal_2018}.)

@@

Accuracy in the diffusion limit is another important property to consider when an IMEX scheme is applied to the two-moment model.  
In the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$, and the two-moment model is approximately governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
In the context of IMEX schemes, the above relationships imply that the following relations should hold~\cite{chu_etal_2018}:
\begin{equation}
   \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=1,\ldots,s,
   \label{diffusionAccuracy}
\end{equation}
where $\vect{e}_{i}$ is the $i$th column of the $s\times s$ identity matrix, $\vect{e}$ is the vector of ones, and $A$ and $\tilde{A}$ are the matrices of the coefficients $(\tilde{a}_{ij})$ and $(a_{ij})$.
Eq.~\eqref{diffusionAccuracy} implies:
\begin{equation}
  c_{i} = \tilde{c}_{i}, \quad i=1,\ldots,s.
\end{equation}
We have proved in \cite{chu_etal_2018} that only IMEX schemes of type~ARS can be both diffusion accurate and convex-invariant.  
(Another short proof follows from the fact that IMEX schemes of type~A have $\tilde{c}_1 = 0$ while $c_i \neq 0$.)

@@

    (Note that the greater the $c_{\Sch}$, the larger the time step can be.
    And $c_{\Sch} \leq 1$.)
    \item Have less than five stages ($s\le4$)\label{cod:statges}.
    \item Are globally stiffly accurate: $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$. 
\end{enumerate}
Fortunately, these IMEX schemes are easy to find.  
(The constraint in \eqref{cod:statges} is introduced from efficiency considerations to limit the number of implicit solves.)
We call the IMEX schemes satisfying the above conditions {PD-ARS} (see also Definition~3 in~\cite{chu_etal_2018}), and we provide two optimal PD-ARS schemes below: PD-ARS2 and PD-ARS3, each limits to the optimal second-order and third-order SSPRK schemes from~\cite{shuOsher_1988}, respectively.
\subsubsection{PD-ARS2}

The optimal 3-statge PD-ARS, PD-ARS2, in the standard double Butcher tableau form, with explicit tableau ($\tilde{A}$) on the left and implicit tableau ($A$) on the right, is given by
\begin{align}
  &\begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0 
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2 & 1/2 \\ \hline
  	  & 0 & 1/2 & 1/2
  \end{array}
\end{align}
Note its explicit tableau is SSPRK2. 
For this scheme, only two implicit solves are needed per time step and $c_{\Sch}= 1$, which implies that the time step restriction for preserving moment realizability is only due to the explicit part.

@@

The optimal 4-stage PD-ARS, PD-ARS3, is given in its standard double Butcher tableau form (explicit tableau on the left and implicit tableau on the right) by
\begin{align}
  &\begin{array}{c | c c c c}
  	    &     &     &     &  \\
  	 1  & 1   &     &     &  \\
  	1/2 & 1/4 & 1/4 &  \\
  	 1  & 1/6 & 1/6 & 2/3 &  \\ \hline
  	    & 1/6 & 1/6 & 2/3 &
  \end{array}
  \qquad
  \begin{array}{c | c c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1/2 & 0 & 1/4 & 1/4 \\ 
  	1 & 0 & 1/6 & 1/6 & 2/3\\\hline
  	  & 0 & 1/6 & 1/6 & 2/3
  \end{array}
\end{align}
Its explicit tableau is SSPRK3. 
This scheme requires three implicit solves per time step, and $c_{\Sch}= 1$.  
Since PD-ARS3 is not more accurate than PD-ARS2 in collision dominated regions (see our results in Section~\ref{se:NumericalTests}), it may not offer any practical advantage over PD-ARS2.

>>>>

Since the set of realizable moments is convex, convex-invariant schemes, which map the initial values into this set, can be used to design realizability-preserving methods for the two-moment model.
Ideally, the scheme should also be high-order accurate and work well in the asymptotic diffusion limit (characterized by frequent collisions and long time scales).  
The following discussion considers the construction of such convex-invariant schemes.

@@

If the IMEX scheme is GSA, $\vect{u}^{n+1} = \vect{u}^{(s)}$.  
Moreover, if $c_{ij},\tilde{c}_{ij}\ge0$ and $a_{ii}>0$, each stage in Eq.~\eqref{eq:imexStagesRewrite} is a convex combination of explicit Euler steps (with time step $\hat{c}_{ij}\dt$), followed by an implicit Euler step.  
Each of the explicit Euler steps has a time step condition that ensures its realizability given by $\hat{c}_{ij}\,\dt\leq\dx$; the CFL condition of the scheme.
Using results proved in~\cite{chu_etal_2018} and discussed in Section~\ref{se:SpatialDiscretization}, the IMEX scheme is convex-invariant and realizability-preserving for the two-moment model in Section~\ref{se:SpatialDiscretization} provided
\begin{equation}
  \max(\hat{c}_{ij})\,\dt \leq \dx.  
\end{equation}
(This CFL condition becomes more restrictive with high-order DG spatial discretization~\cite{chu_etal_2018}.)

@@

Accuracy in the diffusion limit is another important property to consider when an IMEX scheme is applied to the two-moment model.  
In the diffusion limit, the distribution function is nearly isotropic, so $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$, and the two-moment model is approximately governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
In the context of IMEX schemes, the above relationships imply that the following relations should hold~\cite{chu_etal_2018}:
\begin{equation}
   \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=1,\ldots,s,
   \label{diffusionAccuracy}
\end{equation}
where $\vect{e}_{i}$ is the $i$th column of the $s\times s$ identity matrix, $\vect{e}$ is the vector of ones, and $A$ and $\tilde{A}$ are the matrices of the coefficients $(\tilde{a}_{ij})$ and $(a_{ij})$.
Eq.~\eqref{diffusionAccuracy} implies:
\begin{equation}
  c_{i} = \tilde{c}_{i}, \quad i=1,\ldots,s.
\end{equation}
We have proved in \cite{chu_etal_2018} that only IMEX schemes of type~ARS can be both diffusion accurate and convex-invariant.  
(Another short proof follows from the fact that IMEX schemes of type~A have $\tilde{c}_1 = 0$ while $c_i \neq 0$.)

@@

    (Note that the greater the $c_{\Sch}$, the larger the time step can be.
    And $c_{\Sch} \leq 1$.)
    \item Having less than five stages ($s\le4$)\label{cod:statges}.
    \item Are globally stiffly accurate: $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$. 
\end{enumerate}
Fortunately, these IMEX schemes are easy to find.  
(The constraint in \eqref{cod:statges} is introduced from efficiency considerations to limit the number of implicit solves.)
We call the IMEX schemes satisfying the above conditions {PD-ARS} (see also Definition~3 in~\cite{chu_etal_2018}), and we provide two optimal PD-ARS schemes below: PD-ARS2 and PD-ARS3, each limiting to the optimal second-order and third-order SSPRK schemes from~\cite{shuOsher_1988}, respectively.
\subsubsection{PD-ARS2}

The optimal 3-stage PD-ARS, PD-ARS2, in the standard double Butcher tableau form, with explicit tableau ($\tilde{A}$) on the left and implicit tableau ($A$) on the right, is given by
\begin{align}
  &\begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0 
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2 & 1/2 \\ \hline
  	  & 0 & 1/2 & 1/2
  \end{array}
\end{align}
Note its explicit tableau is SSPRK2. 
For this scheme, only two implicit solves are needed per time step and $c_{\Sch}= 1$, which implies that the time step restriction for preserving moment realizability is only due to the explicit part.

@@

The optimal 4-stage PD-ARS, PD-ARS3, is given in its standard double Butcher tableau form (explicit tableau on the left and implicit tableau on the right) by
\begin{align}
  &\begin{array}{c | c c c c}
  	    &     &     &     &  \\
  	 1  & 1   &     &     &  \\
  	1/2 & 1/4 & 1/4 &  \\
  	 1  & 1/6 & 1/6 & 2/3 &  \\ \hline
  	    & 1/6 & 1/6 & 2/3 &
  \end{array}
  \qquad
  \begin{array}{c | c c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1/2 & 0 & 1/4 & 1/4 \\ 
  	1 & 0 & 1/6 & 1/6 & 2/3\\\hline
  	  & 0 & 1/6 & 1/6 & 2/3
  \end{array}
\end{align}
Its explicit tableau is SSPRK3. 
This scheme requires three implicit solves per time step, and $c_{\Sch}= 1$.  
Since PD-ARS3 is not more accurate than PD-ARS2 in collision-dominated regions (see our results in Section~\ref{se:NumericalTests}), it may not offer any practical advantage over PD-ARS2.
